### 2024-04-04 doc/15yrsChange.Rmd
Citation edits & other

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. Although currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
This is problematic in that nuisance species – including non-indigenous insect pests and pathogens, non-indigenous plants, and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.

@@

Another category of nuisance species is mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S. eastern deciduous forests. Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [cote_ecological_2004; @rooney_deer_2001; @brown_future_2000]. White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., XX ) species (REFS). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

In combination, increased canopy tree mortality and reduced canopy species recruitment in the understory is likely to have profound impacts on deciduous forests of eastern North America.  When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a]. In the mid-Atlantic region of the eastern United States, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort when they eventually die [@miller_compounding_2019; @miller_overabundant_2023]. The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp.,[@miller_compounding_2019; @nowacki_climate_2015]), but even these are often not present in sufficient abundance to replace the current canopy (??).

Understanding the combined effects of multiple nuisance species on forest ecosystem processes requires data on both the forest canopy and understory over an extended time. Here, we use 15 years of detailed census from a large forest dynamics plot the mid-Atlantic region of the eastern United States to understand the effects of non-indigenous pests and pathogens and white-tailed deer on the forest carbon balance and regeneration potential.  We test the following hypotheses: (1) canopy tree mortality and associated woody biomass loss are increasing, in large part due to non-indigenous nuisance species, (2) biomass gains from tree growth and recruitment have not kept pace with tree mortality, resulting in a decline in aboveground living biomass (-∆AGB), and (3) deer browsing has decimated the abundance of canopy tree saplings, suggesting that future mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.

@@

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by mcgregor_tree_2021. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021].

@@

As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in @condit_tropical_1998. This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh). The census records the dbh, species, living status, and spatial location of each censused stem. Each individual is marked with an identifying number for sequential data collection in subsequent censuses. In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag. The plot was established in 2008 and recensused three times (2013, 2018, 2023) since, comprising 15 years of forest dynamics data. In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions [@kim_implementing_2022]. Each individual’s location was recorded in the Field Maps application using a digitized map of the plot (citation). In addition to the ForestGEO census, a tree health and mortality survey was conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census [@gonzalez-akre_patterns_2016]. Data were collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of tree health, such as physical damage, potential pathogens and insect infestation.

@@

In addition to the estimation of AWR biomass flux, we analyzed the effects of nuisance species on recruitment through counts of newly recruited stems. We categorized each species as either canopy or understory based on expert knowledge of life history traits. *Quercus* spp. and *Carya* spp. were grouped by genera to better assess differences between the recruitment of canopy and understory species. We then calculated the number of new stems for each census interval for each species or genera. Finally, we assessed the spatial overlap of biomass loss and recruitment of canopy species. We calculated the ∆AGB of canopy trees within each 0.4 hectare quadrat between 2008 and 2023 and the AGB of canopy species in the understory in 2023.

@@

Our investigation supports our hypotheses and demonstrates that the combination of -∆AGB and limited recruitment of canopy species in the understory is compromising the potential carbon sink of this forest, now and into the future. Decreases in AGB were largest in the regions of the plot with high canopy vulnerability to pests and pathogens. These regions also exhibited the lowest recruitment of canopy species in the understory. The region with low deer density had the highest amount of canopy species in the understory. Despite localized increases in AGB, notably in areas with low canopy vulnerability, on net, these increases are not outpacing AGB lost through a combination of increased AWM and decreasing AWG. Moreover, the low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.

@@

# Conclusions /Future Directions

Expand analyses across other temperate forests. Different forest compositions and herbivory dynamics, but many pests and pathogens. Leverage other forest dynamics sites to quantify the impact of nuisance species across large regions.
Investigate productivity responses of surviving canopy. We would expect survivors adjacent to gaps increase productivity as competition decreases, but may be hampered by climate change.
Interactions of global change drivers are critical to understand (though difficult to study).

>>>>

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. Although currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
For instance, one mechanism not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in detrimental ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010]. 
Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
This is problematic in that nuisance species – including non-indigenous insect pests and pathogens, non-indigenous plants, and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.

@@

Another category of nuisance species includes mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S. eastern deciduous forests. Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [cote_ecological_2004; @rooney_deer_2001; @brown_future_2000]. White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., garlic mustard, *Alliaria petiolata*) species (Averill *et al* 2018). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

In combination, increased canopy tree mortality and reduced canopy species recruitment in the understory is likely to have profound impacts on deciduous forests of eastern North America.  When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a]. In the mid-Atlantic region of the eastern United States, forests face a severe regeneration debt, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort of those same species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023]. The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp., [@nowacki_climate_2015], Thompson *et al.* 2013), but even these are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].

Understanding the combined effects of multiple nuisance species on forest ecosystem processes requires data on both the forest canopy and understory over an extended time. Here, we use 15 years of detailed tree census from a large forest dynamics plot the mid-Atlantic region of the eastern United States to understand the effects of non-indigenous pests and pathogens and white-tailed deer on the forest carbon balance and regeneration potential.  We test the following hypotheses: (1) canopy tree mortality and associated woody biomass loss are increasing, in large part due to non-indigenous nuisance species, (2) biomass gains from tree growth and recruitment have not kept pace with tree mortality, resulting in a decline in aboveground living biomass (-∆AGB), and (3) deer browsing has decimated the abundance of canopy tree saplings, suggesting that future tree mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.

@@

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by mcgregor_tree_2021. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].

@@

As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in @condit_tropical_1998. This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh). The census records the dbh, species, living status, and spatial location of each censused stem. Each individual is marked with an identifying number for sequential data collection in subsequent censuses. In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag. The plot was established in 2008 and recensused three times (2013, 2018, 2023) since, comprising 15 years of forest dynamics data. In the last census, systematic data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions [@kim_implementing_2022]. Each individual’s location was recorded in the Field Maps application using a digitized map of the plot (citation). In addition to the ForestGEO census, a tree health and mortality survey has been conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census [@gonzalez-akre_patterns_2016]. Data were collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of tree health, such as physical damage, potential pathogens and insect infestation.

@@

In addition to the estimation of AWR biomass flux, we analyzed the possible effects of nuisance species on recruitment based on counts of newly recruited stems. We categorized each species as either canopy or understory based on expert knowledge of life history traits. *Quercus* spp. and *Carya* spp. were grouped by genera to better assess differences between the recruitment of canopy and understory species. We then calculated the number of new stems for each census interval for each species or genera. Finally, we assessed the spatial overlap of biomass loss and recruitment of canopy species. We calculated the ∆AGB of canopy trees within each 0.4 hectare quadrat between 2008 and 2023 and the AGB of canopy species in the understory in 2023.

@@

Our investigation supports our hypotheses and demonstrates that the combination of -∆AGB and limited recruitment of canopy species in the understory is compromising the potential carbon sink of this forest, now and into the future. Decreases in AGB were largest in the regions of the plot with high canopy vulnerability to pests and pathogens. These regions also exhibited the lowest recruitment of canopy species in the understory. The region with low deer density had the highest amount of canopy species in the understory. Despite localized increases in AGB, notably in areas with low canopy vulnerability, on net, these increases are not outpacing AGB lost through a combination of increased AWM and decreasing AWG. Total ABG stocks in the plot were lower in 2023 than in 2013. Moreover, the low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.

@@

# Future Directions

Understanding interactions between global change drivers, an ongoing area of research, is critical. We would expect that trees unaffected by nuisance species (*e.g.* *Liriodendron tulipifera*, *Quercus rubra*) might see productivity increases as canopy gaps form and competition from nearby trees decreases. However, continuous decreases in AWG throughout the plot suggests that other mechanisms, including increasing maximum temperatures during the growing season (Dow et al 2022), may be hindering tree growth. Tree seedlings of many species, already decimated by deer browsing, are further threatened by rising maximum temperatures (McEwan et al 2011). Increasing mortality within *Quercus* species is another focal area of future research, as species within the *Quercus* genera constitute > 33% of woody productivity at this site (McGregor et al 2020).

Future research should build on our results by expanding these analyses to other temperate forest regions. While other areas will have different forest compositions and herbivory dynamics, increasing pressure from pests and pathogens is pervasive throughout temperate forests (Santini et al. 2012, Ayres and Lombardero 2018). Moreover, while not a large factor at this site, the spread of non-indigenous plant species can further threaten the  regeneration of woody canopy species (Pile Knapp et al, 2023). Large forest dynamics sites (*i.e.* other sites in the ForestGEO network) are well-suited to this research and can be leveraged to quantify how nuisance species compromise present and future forest carbon sequestration. Expanding the scale of these analyses may also facilitate the incorporation of nuisance species into global vegetation and carbon models. Future evaluations of temperate forest health and carbon storage must account for the substantial effects of nuisance species along with other global changes.
